On the properties of discs around accreting brown dwarfs

### 2010-05-18 BD_disc/paper.tex
Changed intro and \cite names, e.g lamm_2004
git-svn-id: https://repository.astro.ex.ac.uk/torus/trunk@2271 25464f21-6099-402c-b30a-640f1d40d38b

% mn2esample.tex
%
% v2.1 released 22nd May 2002 (G. Hutton)
%
% The mnsample.tex file has been amended to highlight
% the proper use of LaTeX2e code with the class file
% and using natbib cross-referencing. These changes
% do not reflect the original paper by A. V. Raveendran.
%
% Previous versions of this sample document were
% compatible with the LaTeX 2.09 style file mn.sty'
% v1.2 released 5th September 1994 (M. Reed)
% v1.1 released 18th July 1994
% v1.0 released 28th January 1994

@@

There is now strong evidence that as brown dwarfs (BDs) form they pass
through a Classical T~Tauri star (CTTS) phase, during which they
possess a flared, dusty circumstellar disc from which they are
actively accreting material (xxx; xxx). The accretion is thought to
proceed via a magnetically-controlled funnel flow mechanism in which
material from a truncated inner disc boundary falls onto the surface
of the star along magnetic field lines
\citep{1990RvMA....3..234C,1991ApJ...370L..39K} xxx.

The truncated inner-edge of the circumstellar disc receives direct
photospheric radiation and is strongly heated. This leads to an
increased scaleheight, or `puffing up' of the inner disc rim, and
subsequent shadowing of the region immediately beyond it (Dullemond
reference). The density-dependent nature of the dust sublimation
temperature (pollack ref) may also play a role in shaping the inner
rim (isella ref; tannirkulam ref).

As an aside, although we recognise that the inner hole in protostellar
discs may be the result of giant planet formation or photoevaporation
(xxx; xxx), these two mechanisms should result in a negligible
mass-accretion rate and therefore would represent a distinct
evolutionary state of brown dwarf disc (BDD) sytems to that of the
CTTS-like phase we are considering here.

The interpretation of the spectral energy distributions (SEDs) of
pre-main-sequence (PMS) stars involves trying to distinguish the
various contributions to the continuum from the hot spots at the base
of the accretion flow, the photospheric flux, and the near-IR flux
from the dusty inner disc. The complexity of the interplay between
these contributions is excerbated by other, geometrical, effects such
as the inclination (which changes the projected area of the inner disc
wall visible to the observer) and the outer disc structure (which may
obsure the inner disc for high inclincations). (xxx; xxx)

In surveys, particularly those attempting to discern the PMS disc
fraction (Xxx), this disentangling usually takes place using broad-band
photometric measures and cuts in colour-colour or colour-magnitude
space (xxx; xxx).  The important quantity here is the separation in
wavelength between the emission peaks for the stellar and thermal disc
components, as it is these components which must be separated. For BDD
systems it is likely that difficulties detecting the disc will be
exacerbated by the lower temperatures of the photosphere and therefore
the smaller separation in wavelength from the thermal disc emission
component, compared to CTTS systems. Therefore, to derive disc
fractions for BD populations we require detailed comparison models
with which to guide disc candidate selection.

Comparison of accretion rates across the pre-main-sequence (PMS) mass
spectrum has indicated that the accretion rate is strongly correlated
with PMS mass,
\citep{2003ApJ...592..266M,2004A&A...424..603N,2006A&A...452..245N},
with an approximate form $\dot{M_{\rm acc}}\propto M_{*}^{~2}$. Since
in the canonical picture the accretion rate is driven by the disc
viscosity, and should be indepedent of the mass of the central object,
this correlation is somewhat surprising. There is some danger that the
correlation is the result of, or at least strengthened by, the
presence of observational biases (xxxx): At the high-mass (CTTS) end of the
mass spectrum the lowest accretion-rates cannot be measured via
continuum methods since the excess is too small in contrast with the
photopsheric emission. The emission lines will also be weak, and
indeed the H$\alpha$ equivalent width (EW) may be less than the
10\AA\ that traditionally demarc?ates classical from weak-lined T Tauri
stars. Such objects should show doppler-broadened profiles, but the
line wings will be weak, and the presence of underlying H$\alpha$
absorption may become dominant. Thus there may be a population of
low-accretion-rate CTTS which are current missing from the surveys.

There are also risks of observational biases at the low-mass end of
the correlation. Radiative-transfer modelling of H$\alpha$ emission
from accreting BDs requires high temperatures ($>$\, 10\,kK) in the
accretion funnels in order to recover the level of emission that is
observed. (Cooling rate arguments can be invoked to explain the
presence of such high temperatures). Recently some continuum
measurements of accreting BDs have been conducted, and these broadly
support the H$\alpha$ rates (xxxx). Nonetheless it is worth considering the
observed effect that mass accretion at typical CTTS rates onto a BD
might have. Parity between the BD photospheric luminosity and the
accretion luminosity occurs at relatively low accretion rates
\citep{2006MNRAS.370L..10C}. Such objects may not be detected as BDs
at all in photometric surveys. Even before such an extreme case is
reached, the additional luminosity provided by the accretion should
have measurable effect on the BD colours. Furthermore the additional
flux will be reprocessed by the disc, altering its scaleheight and
possibly the shape of its inner edge.

Here we present a grid of models of BDD systems, including a
self-consistent treatment of the photospheric and accretion luminosity
sources and the interaction of that flux with the circumstellar disc.
We investigate the impact of the luminosity of the central source on
location and shape of the disc inner rim, as well as the large-scale
structure and flaring of the outer disc. We construct synthetic
colour-colour and colour-magnitude diagrams in order to examine the
efficacy of the photometric selections used to isolate brown dwarfs
and measure disc fractions (xxx;xxx).

@@

In order to model an accreting BDD system we must first model the
underlying photospheric flux. We have adopted a BD stellar interior
and atmospheric model grid and have then constructed the total
photospheric flux for any input value of stellar age and mass by
interpolating for surface gravity ($\log g$), effective temperature
($T_{\rm eff}$), radius ($R_*/R_{\odot}$) and luminosity
($L_*/L_{\odot}$). These values were then used to interpolate
atmospheric spectra for flux (ergs s$^{-1}$cm$^{-2} {\rm \AA}^{-1}$)
from 1200 to 2$\times$ 10$^7$ \AA. The spectra were subsequently
resampled onto 200 logarithmically spaced points. Careful inspection
ensured that no spectral features were removed during resampling. The
stellar interior models used for this study are the `DUSTY00' models
of \cite{2000ApJ...542..464C} combined with the `AMES-Dusty',
atmospheric models of \cite{2000ApJ...542..464C}, which are all
available online
\footnote{http://perso.ens-lyon.fr/france.allard/}. For our $T_{\rm
  eff}$ range of $\approx $3000\,K$<T_{\rm eff}<$1600\,K the
AMES-Dusty atmospheres are the most applicable (2700\,K$>T_{\rm
  eff}>$1700\,K). We did try including dynamic application of
atmospheres based on the derived $T_{\rm eff}$, i.e. using AMES-Cond
for $T_{\rm eff}<$1700\,K, but this resulted in large discontinuities
between the model atmospheres and resulting spectra. Since this only
affects stars at the very edge of our temperature range, i.e. for the
oldest and lowest mass objects (for the AMES-Cond case), we have
adopted the AMES-Dusty models throughout.

@@

In this study we assume that accretion from the central star occurs
along magnetically channelled columns from the inner disc boundary.
For CTTS stars, \cite{2007prpl.conf..479B} show that the magnetic
truncation radius ($R_{\rm mag}$) is less than the co-rotation radius
($R_{\rm co}$), where the angular Keplerian velocity of the
disc is equal to the surface angular velocity of the central star.
Calculations of the magnetic truncation radius depend on derivations
of the surface magnetic field \citep{1991ApJ...370L..39K}. This is
currently unavailable for BD stars, due to increased molecular species
obscuring the Zeeman splitting signatures, that are normally used to
derive stellar surface magnetic fields. Therefore, for our model grid
we have adopted an initial inner disc radius as the co-rotation
radius,
\begin{equation}
{R_{\rm inner}=\left({GM_*\tau^2\over{4\pi ^2}}\right)^{1\over{3}}},
\label{inner_eq}
\end{equation}
where $\tau$ is the stellar rotation period and $R_{\rm inner}$ is the
inner radius. This is effectively adopting a disc-locking mechanism
(without associated angular momentum loss), as for disc-locked stars,
$R_{\rm mag}\approx R_{\rm co}$,
\citep{1991ApJ...370L..39K,1994ApJ...429..781S}. Therefore, for our
model simulations, this inner edge radius is dependent on, and derived
from, the adopted value of the rotational period for the central star,
as well as being weakly dependent on the stellar mass. As discussed in
Section \ref{intro}, the inner disc can be cleared through a number of
mechanisms, including binarity or Giant planet formation,
photoevaporation or photoionisation of the disc and dust grain growth
or settling. For BDD systems where a disc is modeled a treatment of
dust sublimation is included (discussed in Section \ref{dust_edge}).
However, the effects of binarity or Giant planet formation are
neglected. Further to the stellar mass and period required prior to
calculation of the inner disc radius, we require a disc mass (in
stellar masses).

@@

For this work the disc outer edge was set at 300\,AU, this was chosen
as a maximum size of the circumstellar
disc. \cite{2008A&A...486..877B} have shown that the disc outer radius
has little effect on the resulting SED. However, in our subsequent
paper we will include models for outer radii of 100\,AU and plan to
extend this parameter range further to smaller values of the outer
radius in the future.

@@

The combined (accretion plus photosphere) SED is then used as a
boundary condition for the {\sc torus} radiative transfer code and
as a benchmark set of SEDs to model `naked' BD systems. The set of
`naked' photospheres (plus accretion) are diluted by the factor
$(R_*/{\rm distance})^2$ to a distance of 10\,pc. The `negligibly' accreting,
`naked' stars can be used to produce absolute magnitude (and intrinsic
colour) isochrones for comparison. The remainder are used to
model systems showing active accretion where no disc is detected. For
instance, \cite{2009arXiv0901.2603K} find 43 stars within their sample
are actively accreting whilst no disc is detected (out of a total
sample of 1253).

@@

We have used the {\sc torus} radiative transfer code which is
described in \cite{2000MNRAS.315..722H}, including the subsequent
refinements i.e.  addition of Adaptive-Mesh-Refinement (AMR)
introduced in \cite{2004MNRAS.350..565H}. {\sc torus} uses the method
of \cite{1999A&A...344..282L} solve radiative equilibrium. The
simulation also self-consistently solves the equation of
\textit{vertical hydrostatic equilibrium} and dust sublimation for the
disc (described in \cite{2007ApJ...661..374T}).

@@

We have adopted a similar dust model to \cite{2002ApJ...564..887W} 
with the size distribution of dust particle given by,
\begin{equation}
\label{particle_dist}
n(a)da=C_ia^{-q}\times exp^{[-(a/a_c)^p]}da
\end{equation}.
Where $n(a)da$ is the number of particles of size $a$ (within the
increment $da$), $a_{\rm c}$ is the characteristic particle size, with
$p$ and $q$ simply used to control the shape of the
distribution. $C_i$ controls the relative abundance of each
constituent species ($i$) in the dust. \cite{2002ApJ...564..887W}
found that simulated SEDs fit observed SEDs better using this adjusted
size distribution for the dust particles, as opposed to a simple power
law. The best fitting values found in \cite{2002ApJ...564..887W} were
$q=$3.0, $p=$0.6 with $a_{\rm c}=$50\,$\mu$m, also with an associated
maximum and minimum grain size of $a_{\rm min}=$5 nm and $a_{\rm
  max}=$1mm. A slightly steeper power law dependence, with $q=$3.5
(corresponding to the canonical MRN distribution) is more widely
adopted
\citep{2008A&A...486..877B,2008ApJ...676L.143M,2008arXiv0810.2552P}.
Here we have adopted the values of \cite{2002ApJ...564..887W}
except for the parameter $q$ where we have used $q=$3.5.
\cite{2002ApJ...564..887W} calculated the $C_i$ values for amorphous
carbon and silicon by requiring the dust to deplete a solar abundance
of either component completely \citep[using abundances
  from][]{1989GeCoA..53..197A,1993ASPC...40..410N}. We have set
$C_i=$1 and adjusted the species using a grain fractional abundance
\cite[an equivalent process to that of][]{2002ApJ...564..887W},
however we have adjusted these grain fractions using the updated solar
abundances of \cite{2006CoAst.147...76A}. The resulting difference in
opacity between grain fractions matching the work of
\cite{2002ApJ...564..887W} and the new grain fractions is only a
slight enhancement of the silicate feature (due to the relative
abundance of Silicon increasing) which has little effect on the
resulting SEDs. Figure \ref{albedo} shows the resulting albedo, and
scattering and absorption opacities, for our dust population, with a
the vertical dashed line showing 10\,$\mu$m.

@@

As our models assume magnetic-truncation at the co-rotation radius, we
have implicitly generated an inner hole. This also means that the disc
will have an inner wall at this radius. Evidence for inner walls in
circumstellar discs is apparent from the SEDs of disc systems, where a
peak in emission is found between 2 and 3\,$\mu$m. The temperatures
reached by such inner walls are expected to generate thermal flux
contributions within this wavelength range
\citep{2001ApJ...560..957D}.

For low-luminosity systems the dust sublimation radius ($R_{\rm sub}$)
will be coincident with the co-rotation radius. However should
the combined photospheric and accretion luminosities be sufficient the
inner disc will be heated sufficiently that the dust close to the
co-rotation radius will be sublimated, and we must account for this
in our models. Furthermore, since the dust sublimation temperature has
a density dependence \citep{1994ApJ...421..615P}, it is both the location and {\em
  shape} of the inner wall that can change (Isella ref; Tannirkulam ref).

Our treatment of dust sublimation is similar to that detailed in
\cite{2007ApJ...661..374T}, but with some enhancements to ensure a
swift convergence of the sublimated rim. The dust sublimation proceeds
as follows: An initial temperature distribution is found for the
optically-thin limit by setting the global dust-to-gas ratio to a tiny
value. Subsequent radiative equilibrium iterations are performed using
the adopted dust-to-gas ratio of 0.01, but limiting the maximum
optical depth across a given cell to $\tau_{\rm max}$. Cells whose
temperature exceeds the local dust sublimation temperature have their
dust-to-gas ratio set to zero. Radiative equilibrium iterations and
sublimation sweeps are performed at $\tau_{\max} = $0.1, 1, and 10,
with a final iteration of $\tau_{\max} = \infty$. Adequately solving
the radiative-equilibrium necessiates resolving the disc's effective
photosphere, and adaptive mesh refinement is used to split the grid at
the optically-thin/optically-thick boundary in order that the maximum
cell size at this boundary is $<1$ at a wavelength of 5500\,\AA. Once
a self-consistent sublimated rim has been determined the equation of
hydrostatic equilibrium is solved throughout the disc, and the
sublimation iterations are restarted (the change in the density
structure from the hydrostatical equilibrium step naturally feeds back
into the shape of the inner rim). Five hydrostatic equilibrium steps
are performed to ensure convergence, although a stable disc structure
is normally found after three such iterations.

@@

We have integrated using photon-counting and calibrated using a Vega
spectrum for magnitudes in the optical and near-IR regimes. The filter
bandpasses selected are, the optical system of
\cite{1998A&A...333..231B} for \textit{UBVRI} and the \textit{CIT}
system of \cite{1982AJ.....87.1029E,2004PASP..116....9S} for
\textit{JHK}, with the required shift of $-0.015$\,$\mu$m as prescribed by
\cite{2004PASP..116....9S}.

@@

These specific filter sets have been chosen due to their ubiquitous
use and suitability for the derivation of the key stellar parameters
of age, mass and, for populations, disc fractions. Therefore, by
studying the changes of magnitudes (and colours) in these photometric
systems we can test the predicted effects on these derived parameters
caused by changes in the input parameters of our model grid. The
optical magnitudes \textit{VI} are used to explore age related effects
of varying the parameter space. Flux in the \textit{VI} bands is only
minimally affected by accretion flux \citep{1998ASPC..154.1709G} and
disc thermal flux \citep{1998apsf.book.....H} and, additionally, in a
\textit{V, V-I} CMD, the reddening vector lies parallel to the pre-MS,
minimising any age effect of extinction uncertainty, \citep[for a full
discussion see][]{2008MNRAS.386..261M,2007MNRAS.375.1220M}. The
near-IR passbands of \textit{JHK} are most often used to derive
stellar masses as for pre-MS objects the reddening vector, in for
instance, a \textit{J, J-K} CMD, is almost perpendicular to the
isochrones, minimising the mass effect of extinction uncertainties
\citep[the \textit{CIT} systems was chosen to
match][]{2000ApJ...542..464C}. Finally, as is now well documented the
\textit{Spitzer} IRAC passbands provide the best data with which to
unambiguously separate naked and star-disc systems
\citep{2005ApJ...620L..51L}. In addition, at longer wavelengths, the
MIPS instrument provides sensitivity to disc systems at much larger
radii (or debris discs).

@@

Typical disc lifetimes for solar type stars are of order 10\,Myrs
\citep{2001ApJ...553L.153H}. Therefore, we have adopted input ages of
1 and 10\,Myrs for our model grid, to span the approximate range of
ages over which the discs influence will be important.

@@

Data for rotation rates, from periodic variability surveys, are widely
available for a range of different age clusters of TTS. However, fewer
studies exist on the rotation rates of BD mass objects. To select a
feasible range of periods for our model grid we have collated the
results from some recent studies of BD variability. Rotation period
data for $\sigma$ Ori, at an age of $\approx$ 3 Myrs
\citep{2008MNRAS.386..261M}, is studied in \cite{2004A&A...419..249S},
where periods are found over the range 5.78$-$74.4 hours ($\approx$
0.24$-$3.1 days) for BD mass objects. In some cases the periodic
variability is irregular and assumed to come from active accretion hot
spots on the BD surface \citep[see][for discussion of variability
causes]{1995A&A...299...89B,2007prpl.conf..297H}. Infrared excesses
and evidence of accretion are also found to correlate with slower
rotating TTS, which the authors construe as evidence for a disc
locking scenario. \cite{2004A&A...419..249S} also find a $mass\propto
period$ relationship extending into the BD mass regime, this faster
rotation in younger mass objects is taken as evidence for a decreasing
effectiveness of the angular momentum removal of the disc locking
mechanism, i.e. imperfect disc locking. In addition,
\cite{2004A&A...419..249S} argue that when compared with data from
younger clusters disc lifetimes can be seen to decrease with central
object mass.

>>>>

% mn2esample.tex
%
% v2.1 released 22nd May 2002 (G. Hutton)
%
% The mnsample.tex file has been amended to highlight
% the proper use of LaTeX2e code with the class file
% and using natbib cross-referencing. These changes
% do not reflect the original paper by A. V. Raveendran.
%
% Previous versions of this sample document were
% compatible with the LaTeX 2.09 style file mn.sty'
% v1.2 released 5th September 1994 (M. Reed)
% v1.1 released 18th July 
% v1.0 released 28th January 1994

@@

There is now strong evidence that as brown dwarfs (BDs) form they pass
through a Classical T~Tauri star (CTTS) phase, during which they
possess a flared, dusty circumstellar disc from which they are
actively accreting material
\citep{jayawardhana_2003,mohanty_2004}. The accretion is
thought to proceed via a magnetically-controlled funnel flow mechanism
in which material from a truncated inner disc boundary falls onto the
surface of the star along magnetic field lines
\citep{camenzind_1990,koenigl_1991,mohanty_2008}.

The truncated inner-edge of the circumstellar disc receives direct
photospheric radiation and is strongly heated. This leads to an
increased scaleheight, or `puffing up' of the inner disc rim, and
subsequent shadowing of the region immediately beyond it
\citep{dullemond_2001}. The density-dependent nature of the dust
sublimation temperature \citep{pollack_1994} may also play a
role in shaping the inner rim
\citep{isella_2005,tannirkulam_2007}.

As an aside, although we recognise that the inner hole in protostellar
discs may be the result of giant planet formation or photoevaporation
\citep{dahm_2009,najita_2007}, these two mechanisms
should result in a negligible mass-accretion rate
\citep{dahm_2009} and therefore would represent a distinct
evolutionary state of brown dwarf disc (BDD) sytems to that of the
CTTS-like phase we are considering here.

The interpretation of the spectral energy distributions (SEDs) of
pre-main-sequence (PMS) stars involves trying to distinguish the
various contributions to the continuum from the hot spots at the base
of the accretion flow, the photospheric flux, and the near-IR flux
from the dusty inner disc. The complexity of the interplay between
these contributions is excerbated by other, geometrical, effects such
as the inclination (which changes the projected area of the inner disc
wall visible to the observer) and the outer disc structure (which may
obsure the inner disc for high inclincations)
\citep{walker_2004,tannirkulam_2007}.

In surveys, particularly those attempting to discern the PMS disc
fraction , this disentangling usually takes place using broad-band
photometric measures and cuts in colour-colour or colour-magnitude
space
\citep{luhman_2005,Luhman_2008,gutermuth_2008}.
The important quantity here is the separation in wavelength between
the emission peaks for the stellar and thermal disc components, as it
is these components which must be separated. For BDD systems it is
likely that difficulties detecting the disc will be exacerbated by the
lower temperatures of the photosphere and therefore the smaller
separation in wavelength from the thermal disc emission component,
compared to CTTS systems. Therefore, to derive disc fractions for BD
populations we require detailed comparison models with which to guide
disc candidate selection.

Comparison of accretion rates across the pre-main-sequence (PMS) mass
spectrum has indicated that the accretion rate is strongly correlated
with PMS mass,
\citep{muzerolle_2003,natta_2004,natta_2006},
with an approximate form $\dot{M_{\rm acc}}\propto M_{*}^{~2}$. Since
in the canonical picture the accretion rate is driven by the disc
viscosity, and should be indepedent of the mass of the central object,
this correlation is somewhat surprising. There is some danger that the
correlation is the result of, or at least strengthened by, the
presence of observational biases \citep{clarke_2006}: At the
high-mass (CTTS) end of the mass spectrum the lowest accretion-rates
cannot be measured via continuum methods since the excess is too small
in contrast with the photopsheric emission. The emission lines will
also be weak, and indeed the H$\alpha$ equivalent width (EW) may be
less than the 10\AA\ that traditionally demarcates classical from
weak-lined T Tauri stars. Such objects should show doppler-broadened
profiles, but the line wings will be weak, and the presence of
underlying H$\alpha$ absorption may become dominant. Thus there may be
a population of low-accretion-rate CTTS which are current missing from
the surveys.

There are also risks of observational biases at the low-mass end of
the correlation. Radiative-transfer modelling of H$\alpha$ emission
from accreting BDs requires high temperatures ($>$\, 10\,kK) in the
accretion funnels in order to recover the level of emission that is
observed. (Cooling rate arguments can be invoked to explain the
presence of such high temperatures). Recently some continuum
measurements of accreting BDs have been conducted, and these broadly
support the H$\alpha$ rates
\citep{muzerolle_2003,natta_2004,natta_2006}. Nonetheless
it is worth considering the observed effect that mass accretion at
typical CTTS rates onto a BD might have. Parity between the BD
photospheric luminosity and the accretion luminosity occurs at
relatively low accretion rates \citep{clarke_2006}. Such
objects may not be detected as BDs at all in photometric surveys. Even
before such an extreme case is reached, the additional luminosity
provided by the accretion should have measurable effect on the BD
colours. Furthermore the additional flux will be reprocessed by the
disc, altering its scaleheight and possibly the shape of its inner
edge.

Here we present a grid of models of BDD systems, including a
self-consistent treatment of the photospheric and accretion luminosity
sources and the interaction of that flux with the circumstellar disc.
We investigate the impact of the luminosity of the central source on
location and shape of the disc inner rim, as well as the large-scale
structure and flaring of the outer disc. We construct synthetic
colour-colour and colour-magnitude diagrams in order to examine the
efficacy of the photometric selections used to isolate brown dwarfs
and measure disc fractions
\citep{luhman_2005,Luhman_2008,gutermuth_2008}.

@@

In order to model an accreting BDD system we must first model the
underlying photospheric flux. We have adopted a BD stellar interior
and atmospheric model grid and have then constructed the total
photospheric flux for any input value of stellar age and mass by
interpolating for surface gravity ($\log g$), effective temperature
($T_{\rm eff}$), radius ($R_*/R_{\odot}$) and luminosity
($L_*/L_{\odot}$). These values were then used to interpolate
atmospheric spectra for flux (ergs s$^{-1}$cm$^{-2} {\rm \AA}^{-1}$)
from 1200 to 2$\times$ 10$^7$ \AA. The spectra were subsequently
resampled onto 200 logarithmically spaced points. Careful inspection
ensured that no spectral features were removed during resampling. The
stellar interior models used for this study are the `DUSTY00' models
of \cite{chabrier_2000} combined with the `AMES-Dusty',
atmospheric models of \cite{chabrier_2000}, which are all
available online
\footnote{http://perso.ens-lyon.fr/france.allard/}. For our $T_{\rm
  eff}$ range of $\approx $3000\,K$<T_{\rm eff}<$1600\,K the
AMES-Dusty atmospheres are the most applicable (2700\,K$>T_{\rm
  eff}>$1700\,K). We did try including dynamic application of
atmospheres based on the derived $T_{\rm eff}$, i.e. using AMES-Cond
for $T_{\rm eff}<$1700\,K, but this resulted in large discontinuities
between the model atmospheres and resulting spectra. Since this only
affects stars at the very edge of our temperature range, i.e. for the
oldest and lowest mass objects (for the AMES-Cond case), we have
adopted the AMES-Dusty models throughout.

@@

In this study we assume that accretion from the central star occurs
along magnetically channelled columns from the inner disc boundary.
For CTTS stars, \cite{bouvier_2007} show that the magnetic
truncation radius ($R_{\rm mag}$) is less than the co-rotation radius
($R_{\rm co}$), where the angular Keplerian velocity of the
disc is equal to the surface angular velocity of the central star.
Calculations of the magnetic truncation radius depend on derivations
of the surface magnetic field \citep{koenigl_1991}. This is
currently unavailable for BD stars, due to increased molecular species
obscuring the Zeeman splitting signatures, that are normally used to
derive stellar surface magnetic fields. Therefore, for our model grid
we have adopted an initial inner disc radius as the co-rotation
radius,
\begin{equation}
{R_{\rm inner}=\left({GM_*\tau^2\over{4\pi ^2}}\right)^{1\over{3}}},
\label{inner_eq}
\end{equation}
where $\tau$ is the stellar rotation period and $R_{\rm inner}$ is the
inner radius. This is effectively adopting a disc-locking mechanism
(without associated angular momentum loss), as for disc-locked stars,
$R_{\rm mag}\approx R_{\rm co}$,
\citep{koenigl_1991,shu_1994}. Therefore, for our
model simulations, this inner edge radius is dependent on, and derived
from, the adopted value of the rotational period for the central star,
as well as being weakly dependent on the stellar mass. As discussed in
Section \ref{intro}, the inner disc can be cleared through a number of
mechanisms, including binarity or Giant planet formation,
photoevaporation or photoionisation of the disc and dust grain growth
or settling. For BDD systems where a disc is modeled a treatment of
dust sublimation is included (discussed in Section \ref{dust_edge}).
However, the effects of binarity or Giant planet formation are
neglected. Further to the stellar mass and period required prior to
calculation of the inner disc radius, we require a disc mass (in
stellar masses).

@@

For this work the disc outer edge was set at 300\,AU, this was chosen
as a maximum size of the circumstellar
disc. \cite{bouy_2008} have shown that the disc outer radius
has little effect on the resulting SED. However, in our subsequent
paper we will include models for outer radii of 100\,AU and plan to
extend this parameter range further to smaller values of the outer
radius in the future.

@@

The combined (accretion plus photosphere) SED is then used as a
boundary condition for the {\sc torus} radiative transfer code and
as a benchmark set of SEDs to model `naked' BD systems. The set of
`naked' photospheres (plus accretion) are diluted by the factor
$(R_*/{\rm distance})^2$ to a distance of 10\,pc. The `negligibly' accreting,
`naked' stars can be used to produce absolute magnitude (and intrinsic
colour) isochrones for comparison. The remainder are used to
model systems showing active accretion where no disc is detected. For
instance, \cite{kennedy_2009} find 43 stars within their sample
are actively accreting whilst no disc is detected (out of a total
sample of 1253).

@@

We have used the {\sc torus} radiative transfer code which is
described in \cite{harries_2000}, including the subsequent
refinements i.e.  addition of Adaptive-Mesh-Refinement (AMR)
introduced in \cite{harries_2004}. {\sc torus} uses the method
of \cite{lucy_1999} solve radiative equilibrium. The
simulation also self-consistently solves the equation of
\textit{vertical hydrostatic equilibrium} and dust sublimation for the
disc (described in \cite{tannirkulam_2007}).

@@

We have adopted a similar dust model to \cite{wood_2002} 
with the size distribution of dust particle given by,
\begin{equation}
\label{particle_dist}
n(a)da=C_ia^{-q}\times exp^{[-(a/a_c)^p]}da
\end{equation}.
Where $n(a)da$ is the number of particles of size $a$ (within the
increment $da$), $a_{\rm c}$ is the characteristic particle size, with
$p$ and $q$ simply used to control the shape of the
distribution. $C_i$ controls the relative abundance of each
constituent species ($i$) in the dust. \cite{wood_2002}
found that simulated SEDs fit observed SEDs better using this adjusted
size distribution for the dust particles, as opposed to a simple power
law. The best fitting values found in \cite{wood_2002} were
$q=$3.0, $p=$0.6 with $a_{\rm c}=$50\,$\mu$m, also with an associated
maximum and minimum grain size of $a_{\rm min}=$5 nm and $a_{\rm
  max}=$1mm. A slightly steeper power law dependence, with $q=$3.5
(corresponding to the canonical MRN distribution) is more widely
adopted
\citep{bouy_2008,morrow_2008,pascucci_2008}.
Here we have adopted the values of \cite{wood_2002}
except for the parameter $q$ where we have used $q=$3.5.
\cite{wood_2002} calculated the $C_i$ values for amorphous
carbon and silicon by requiring the dust to deplete a solar abundance
of either component completely \citep[using abundances
  from][]{anders_1989,noels_1993}. We have set
$C_i=$1 and adjusted the species using a grain fractional abundance
\cite[an equivalent process to that of][]{wood_2002},
however we have adjusted these grain fractions using the updated solar
abundances of \cite{asplund_2006}. The resulting difference in
opacity between grain fractions matching the work of
\cite{wood_2002} and the new grain fractions is only a
slight enhancement of the silicate feature (due to the relative
abundance of Silicon increasing) which has little effect on the
resulting SEDs. Figure \ref{albedo} shows the resulting albedo, and
scattering and absorption opacities, for our dust population, with a
the vertical dashed line showing 10\,$\mu$m.

@@

As our models assume magnetic-truncation at the co-rotation radius, we
have implicitly generated an inner hole. This also means that the disc
will have an inner wall at this radius. Evidence for inner walls in
circumstellar discs is apparent from the SEDs of disc systems, where a
peak in emission is found between 2 and 3\,$\mu$m. The temperatures
reached by such inner walls are expected to generate thermal flux
contributions within this wavelength range
\citep{dullemond_2001}.

For low-luminosity systems the dust sublimation radius ($R_{\rm sub}$)
will be coincident with the co-rotation radius. However should
the combined photospheric and accretion luminosities be sufficient the
inner disc will be heated sufficiently that the dust close to the
co-rotation radius will be sublimated, and we must account for this
in our models. Furthermore, since the dust sublimation temperature has
a density dependence \citep{pollack_1994}, it is both the location and {\em
  shape} of the inner wall that can change (Isella ref; Tannirkulam ref).

Our treatment of dust sublimation is similar to that detailed in
\cite{tannirkulam_2007}, but with some enhancements to ensure a
swift convergence of the sublimated rim. The dust sublimation proceeds
as follows: An initial temperature distribution is found for the
optically-thin limit by setting the global dust-to-gas ratio to a tiny
value. Subsequent radiative equilibrium iterations are performed using
the adopted dust-to-gas ratio of 0.01, but limiting the maximum
optical depth across a given cell to $\tau_{\rm max}$. Cells whose
temperature exceeds the local dust sublimation temperature have their
dust-to-gas ratio set to zero. Radiative equilibrium iterations and
sublimation sweeps are performed at $\tau_{\max} = $0.1, 1, and 10,
with a final iteration of $\tau_{\max} = \infty$. Adequately solving
the radiative-equilibrium necessiates resolving the disc's effective
photosphere, and adaptive mesh refinement is used to split the grid at
the optically-thin/optically-thick boundary in order that the maximum
cell size at this boundary is $<1$ at a wavelength of 5500\,\AA. Once
a self-consistent sublimated rim has been determined the equation of
hydrostatic equilibrium is solved throughout the disc, and the
sublimation iterations are restarted (the change in the density
structure from the hydrostatical equilibrium step naturally feeds back
into the shape of the inner rim). Five hydrostatic equilibrium steps
are performed to ensure convergence, although a stable disc structure
is normally found after three such iterations.

@@

We have integrated using photon-counting and calibrated using a Vega
spectrum for magnitudes in the optical and near-IR regimes. The filter
bandpasses selected are, the optical system of
\cite{bessell_1998} for \textit{UBVRI} and the \textit{CIT}
system of \cite{elias_1982,stephens_2004} for
\textit{JHK}, with the required shift of $-0.015$\,$\mu$m as prescribed by
\cite{stephens_2004}.

@@

These specific filter sets have been chosen due to their ubiquitous
use and suitability for the derivation of the key stellar parameters
of age, mass and, for populations, disc fractions. Therefore, by
studying the changes of magnitudes (and colours) in these photometric
systems we can test the predicted effects on these derived parameters
caused by changes in the input parameters of our model grid. The
optical magnitudes \textit{VI} are used to explore age related effects
of varying the parameter space. Flux in the \textit{VI} bands is only
minimally affected by accretion flux \citep{gullbring_1998} and
disc thermal flux \citep{hartmann_1998} and, additionally, in a
\textit{V, V-I} CMD, the reddening vector lies parallel to the pre-MS,
minimising any age effect of extinction uncertainty, \citep[for a full
discussion see][]{mayne_2008,mayne_2007}. The
near-IR passbands of \textit{JHK} are most often used to derive
stellar masses as for pre-MS objects the reddening vector, in for
instance, a \textit{J, J-K} CMD, is almost perpendicular to the
isochrones, minimising the mass effect of extinction uncertainties
\citep[the \textit{CIT} systems was chosen to
match][]{chabrier_2000}. Finally, as is now well documented the
\textit{Spitzer} IRAC passbands provide the best data with which to
unambiguously separate naked and star-disc systems
\citep{luhman_2005}. In addition, at longer wavelengths, the
MIPS instrument provides sensitivity to disc systems at much larger
radii (or debris discs).

@@

Typical disc lifetimes for solar type stars are of order 10\,Myrs
\citep{haisch_2001}. Therefore, we have adopted input ages of
1 and 10\,Myrs for our model grid, to span the approximate range of
ages over which the discs influence will be important.

@@

Data for rotation rates, from periodic variability surveys, are widely
available for a range of different age clusters of TTS. However, fewer
studies exist on the rotation rates of BD mass objects. To select a
feasible range of periods for our model grid we have collated the
results from some recent studies of BD variability. Rotation period
data for $\sigma$ Ori, at an age of $\approx$ 3 Myrs
\citep{mayne_2008}, is studied in \cite{scholz_2004},
where periods are found over the range 5.78$-$74.4 hours ($\approx$
0.24$-$3.1 days) for BD mass objects. In some cases the periodic
variability is irregular and assumed to come from active accretion hot
spots on the BD surface \citep[see][for discussion of variability
causes]{bouvier_1995,herbst_2007}. Infrared excesses
and evidence of accretion are also found to correlate with slower
rotating TTS, which the authors construe as evidence for a disc
locking scenario. \cite{scholz_2004} also find a $mass\propto
period$ relationship extending into the BD mass regime, this faster
rotation in younger mass objects is taken as evidence for a decreasing
effectiveness of the angular momentum removal of the disc locking
mechanism, i.e. imperfect disc locking. In addition,
\cite{scholz_2004} argue that when compared with data from
younger clusters disc lifetimes can be seen to decrease with central
object mass.
